An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-22 comment.tex
push ot alale

\affil[1]{Porter School of the Environment and Earth Sciences, The Faculty of Exact Sciences, Tel Aviv
  University\\ Tel Aviv, Israel}
\affil[2]{School of Public Health, The Faculty of Medicine, Tel Aviv University\\ Tel Aviv, Israel}
\affil[3]{The Gertner Institute for Epidemiology and Health Policy Research, Tel Hashomer, Israel}
%\email{yair.daon@gmail.com}
\date{}

\maketitle
%%% MAKE TITLE PAGE %%%
%% \bigskip
%% \bigskip
%% \bigskip
%% \bigskip
%% \begin{itemize}
%%     \item Article type: Brief report.
%%     \item Corresponding author: Uri Obolski, email uriobols@tauex.tau.ac.il, Phone +972526309508
%%     \item Running head: Inflated false-negative rates in pooled SARS-CoV-2 tests.
%%     \item Conflict of interests: None.
%%     \item Financial support: Yair Daon was supported by a
%%     post-doctoral fellowship from the Tel Aviv University Center for
%%     Combating Pandemics and the Raymond and Beverly Sackler Dean's
%%     Post-Doctoral fellowship.
%%     \item Data is presented in the manuscript itself and can be found in a paper we reference to. Calculations are simple and can be easily implemented in any programming language.
%% \end{itemize}
%% \newpage

@@

\section{Methods}
Before delving into our probabilistic assumptions, we briefly explain
the process of testing for SARS-CoV-2 by RT-PCR in
Section~\ref{subsec:rtpcr}. Our exposition intentionally avoids many
details, and a comprehensive review of RT-PCR can be found in
\cite{COVID-RTPCR, PCRBook}. We then present the \emph{common
  assumption} in Section~\ref{subsec:common} and refute it in
Section~\ref{subsec:refute}. Our model is presented in
Section~\ref{subsec:ours}.

\subsection{RT-PCR}\label{subsec:rtpcr}
PCR is a process in which a targeted DNA strand's frequency is
amplified to create billions of copies in a reaction mixture. At the
end of this \emph{amplification} process, the presence of targeted DNA
molecules in a reaction mixture can be confidently
determined. Arguably, the best way to understand PCR is as a chain
reaction: Each targeted DNA molecule is copied into two, that are
copied into four and so forth. This cascade gives PCR its name:
polymerase chain reaction.

@@

One possible source of failed detection of SARS-CoV-2 RNA in pooled
RT-PCR testing (a \emph{false-negative}) is sample dilution. When
pooling, several samples are mixed, so the concentration of viral RNA
is reduced. This effect may cause a delay in amplification
\cite{DorfmanYuvalDor}, no detection and, consequently, a
false-negative. Another potential source for false-negatives is a
failed sampling process.

@@

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N} \Pr(\theta_2,\dots,\theta_N) \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)\Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}\left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}, \text{ by Newton's Binomial Theorem.} \\
  \end{split}
\end{align}

@@

\section{Results}\label{section:results}
We plot $\rel{ours}$ for varying prevalence $q$ and sensitivity $\Se$
values. As recommended by \cite{DorfmanYuvalDor}, we apply different
pool sizes $N$, for different prevalence values. We observe that for a
false-positive rate $\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of
reasonable sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\rel{ours}$ can be expected (Figure \ref{figy}). Interestingly, an
increase in infection prevalence monotonically decreases the scheme
false-negative rate, as can also be easily seen from equation
\eqref{eq:sfn}. For the chosen parameter ranges, the increase in the
single test false-negative rates increases the relative error
$\rel{ours}$. These effects can be seen in Figure \ref{figy} (left panel),
upon conditioning on pool size. Extending the range for $\Sp$ yields
no qualitative differences. We further compare $\rel{ours}$ to
$\rel{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{figy}, right
panel).
\begin{figure}[H]
    \centering
    \includegraphics[width=18cm]{heatmap.jpg}
    \caption{Relative increase in Dorfman pooling false-negative rates
      $\rel{ours}$. Left: Colors represent $\rel{ours}$, the relative
      percentage increase in the scheme false-negative rates relative
      to the single test false-negative rates (eq
      \eqref{eq:erel}). Right: colors represent the difference between
      $\rel{common}$ and $\rel{ours}$. The disease prevalence $q$, is
      varied on the x-axis, while the test sensitivity is varied on
      the y-axis. Pool size $N$, was chosen according to $q$ as in
      \cite{DorfmanYuvalDor}.}\label{figy}
 \end{figure}

\section{Discussion}
Although Dorfman pooling improves testing throughput, we have shown
that it can increase $\rel{ours}$ --- the false-negative rates relative to
individual testing. Furthermore, low values of infection prevalence,
or low values of single test false-negative rates, increase
$\rel{ours}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{figy}).

Our results lead to a fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when prevalence is low, such circumstances are exactly those leading
to a substantial increase in false-negative rates.

%% An approximation to our results has been previously considered: $\mi
%% \approx 1-\Se^2 $, see \cite{Simplistic1, Simplistic2, Kim,
%%   OptimalDorfmanPool}. Such an approximation arises when one assumes
%% that the probability of a false-negative is identical for the pooled
%% and single tests. Although simplistic in nature, this approximation
%% does capture the intuition behind our results: For Donald to be
%% considered negative under Dorfman pooling, he has to test negative
%% twice. In addition to this approximation only being an upper bound
%% \cite{Simplistic2}, it does not account for the effect of infection
%% prevalence on the false-negative rates under Dorfman
%% pooling. Supporting the importance of this effect, such an association
%% between prevalence and the scheme false-negative rate under Dorfman
%% pooling has been empirically noted \cite{DorfmanYuvalDor}.

%% Although we have shown the inherent risk of Dorfman pooling, this
%% shortcoming applies to other pooling schemes. Pooling schemes
%% (e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
%% positive pooled results to correctly identify Donald as
%% infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
%% If the first pool yields a positive result, it is split in two. Then
%% the splitting is repeated until resulting pools are negative or
%% individuals are tested separately. With an initial pool size of 32,
%% Donald will necessarily have to test positive in pools of size 32, 16,
%% 8, 4 and 2, as well as in a single test, for the scheme to correctly
%% identify him as infected. Compare this to the Dorfman scheme that
%% requires a positive test in a pool of size $N=8$, and an additional
%% single positive test to identify Donald as infected. The hierarchical
%% pooling scheme of \cite{Lion, Kim} will necessarily yield more
%% false-negatives than Dorfman pooling --- there are additional places
%% for it to fail.

>>>>

\affil[1]{Porter School of the Environment and Earth Sciences, The
  Faculty of Exact Sciences, Tel Aviv University\\ Tel Aviv, Israel}

\affil[2]{School of Public Health, The Faculty of Medicine, Tel Aviv
  University\\ Tel Aviv, Israel}

\affil[3]{The Gertner Institute for Epidemiology and Health Policy
  Research, Tel Hashomer, Israel}
%\email{yair.daon@gmail.com}
\date{}

\maketitle

@@

\subsection{RT-PCR}\label{subsec:rtpcr}
We briefly explain the process of testing for SARS-CoV-2 by RT-PCR in
Section~\ref{subsec:rtpcr}. Our exposition intentionally avoids many
details, and a comprehensive review of RT-PCR can be found in
\cite{COVID-RTPCR, PCRBook}.

PCR is a process in which a targeted DNA strand's frequency is
amplified to create billions of copies in a reaction mixture. At the
end of this \emph{amplification} process, the presence of targeted DNA
molecules in a reaction mixture can be confidently
determined. Arguably, the best way to understand PCR is as a chain
reaction: Each targeted DNA molecule is copied into two, that are
copied into four and so forth. This cascade gives PCR its name:
polymerase chain reaction.

@@

\subsection{Pooling}
One possible source of failed detection of SARS-CoV-2 RNA in pooled
RT-PCR testing (a \emph{false-negative}) is sample dilution. When
pooling, several samples are mixed, so the concentration of viral RNA
is reduced. This effect may cause a delay in amplification
\cite{DorfmanYuvalDor}, no detection and, consequently, a
false-negative. Another potential source for false-negatives is a
failed sampling process.

@@

\section{Methods}
We present the \emph{common assumption} in Section~\ref{subsec:common}
and refute it in Section~\ref{subsec:refute}. Our model is presented
in Section~\ref{subsec:ours}.

@@

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N}
    \Pr(\theta_2,\dots,\theta_N)
    \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)
    \Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}
    \left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}, \text{ by Newton's Binomial Theorem.} \\
  \end{split}
\end{align}

@@

\section{Results}\label{section:results}
We plot $\rel{ours}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\rel{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\rel{ours}$ can be expected (Figure \ref{figy}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\rel{ours}$. These effects can be seen in Figure \ref{figy} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\rel{ours}$ to
$\rel{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{figy}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\rel{ours}$. Left: Colors represent $\rel{ours}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\rel{common}$ and $\rel{ours}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

\section{Discussion}
We have presented a novel model for pooled tests. It is hard to
assess the validity of our model, since this would require
considerably more data than available in \cite{Salazar}. However, our
model is as simple as possible, given that the standard assumptions
are refuted. We have no doubt that further study of the assumptions
underlying pooling is due. Furthermore, adapting common group testing
algorithms \cite{Kim} to novel error models should be further
studied.

Our revised model has some implications for the practice of
pooling. For example, although Dorfman pooling improves testing
throughput, we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\rel{ours}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{figy}).

Our results lead to a fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when prevalence is low, such circumstances are exactly those leading
to a substantial increase in false-negative rates. Such behavior of
pooling is not implied by the common assumption

%% The common assumption implies that one assumes that the probability of
%% a false-negative is identical for the pooled and single
%% tests. Although simplistic in nature, the common assumption does capture
%% the intuition behind our results: For Donald to be considered negative
%% under Dorfman pooling, he has to test negative twice. In addition to
%% this approximation only being an upper bound \cite{Simplistic2}, it
%% does not account for the effect of infection prevalence on the
%% false-negative rates under Dorfman pooling. Supporting the importance
%% of this effect, such an association between prevalence and the scheme
%% false-negative rate under Dorfman pooling has been empirically noted
%% \cite{DorfmanYuvalDor}.

Although our application focused on Dorfman pooling, our analysis
applies to other pooling schemes. Pooling schemes
(e.g. \cite{MatrixPooling,Lion, Kim}), require some sequence of
positive pooled results to correctly identify Donald as
infected. Consider the hierarchical pooling scheme \cite{Lion, Kim}:
If the first pool yields a positive result, it is split in two. Then
the splitting is repeated until resulting pools are negative or
individuals are tested separately. With an initial pool size of 32,
Donald will necessarily have to test positive in pools of size 32, 16,
8, 4 and 2, as well as in a single test, for the scheme to correctly
identify him as infected. Compare this to the Dorfman scheme that
requires a positive test in a pool of size $N=8$, and an additional
single positive test to identify Donald as infected. The hierarchical
pooling scheme of \cite{Lion, Kim} will necessarily yield more
false-negatives than Dorfman pooling --- there are additional places
for it to fail.
